The low-frequency environment of the Murchison Widefield Array: radio-frequency interference analysis and mitigation

### 2014-11-11 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Paul Hancock's comments

\author[A.~R.~Offringa et al.]{A.~R.~Offringa\ANU$^,$\CAASTRO$^,$\ASTRON\thanks{Corresponding author. E-mail: \url{offringa@gmail.com}},
R.~B.~Wayth\Curtin$^,$\CAASTRO,
N.~Hurley-Walker\Curtin,
D.~L.~Kaplan\UWisc,
N.~Barry\UW,
A.~P.~Beardsley\UW,
M.~E.~Bell\CASS,
G.~Bernardi\SKASA,
J.~D.~Bowman\ASU, 
F.~Briggs\ANU,
J.~R.~Callingham\USydney$^,$\CAASTRO,
R.~J.~Cappallo\Haystack, 
P.~Carroll\UW,
%A.~de~Oliveira-Costa
A.~A.~Deshpande\RRI, 
J.~S.~Dillon\MIT,
K.~S.~Dwarakanath\RRI,
A.~Ewall-Wice\MIT,
L.~Feng\MIT,
B.-Q.~For\UWA,
B.~M.~Gaensler\USydney$^,$\CAASTRO, 
L.~J.~Greenhill\CfA,
P.~Hancock\Curtin,
K.~Han-Seek\UMelbourne,
B.~J.~Hazelton\UW,
J.~N.~Hewitt\MIT,
L.~Hindson\Victoria,
D.~C.~Jacobs\ASU, 
M.~Johnston-Hollitt\Victoria,
A.~Kapinska\UWA$^,$\CAASTRO,
P.~Kittiwisit\ASU,
E.~Lenc\USydney$^,$\CAASTRO,
J.~Line\UMelbourne,
A.~Loeb\CfA,
C.~J.~Lonsdale\Haystack, 
B.~McKinley\ANU$^,$\CAASTRO,
S.~R.~McWhirter\Haystack,
D.~A.~Mitchell\CASS$^,$\CAASTRO, 
M.~F.~Morales\UW, 
E.~Morgan\MIT, 
J.~Morgan\Curtin,
A.~R.~Neben\MIT,
D.~Oberoi\Tata, 
S.~M.~Ord\Curtin$^,$\CAASTRO,
S.~Paul\RRI,
B.~Pindor\UMelbourne,
J.~C.~Pober\UW,
T.~Prabu\RRI, 
P.~Procopio\UMelbourne,
J.~Riding\UMelbourne,  
N.~Udaya~Shankar\RRI, 
S.~Sethi\RRI,
K.~S.~Srivani\RRI,
L.~Staveley-Smith\UWA,
R.~Subrahmanyan\RRI$^,$\CAASTRO, 
I.~S.~Sullivan\UW,
M.~Tegmark\MIT,
N.~Thyagarajan\ASU,
S.~J.~Tingay\Curtin$^,$\CAASTRO, 
C.~M.~Trott\Curtin$^,$\CAASTRO,
R.~L.~Webster\UMelbourne$^,$\CAASTRO, 
A.~Williams\Curtin, 
C.~L.~Williams\MIT,
C.~Wu\UWA,
J.~S.~Wyithe\UMelbourne,
Q. Zheng\Victoria
\\
\ANU{}Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
\ASTRON{}Netherlands Institute for Radio Astronomy (ASTRON), PO Box 2, 7990 AA Dwingeloo, The Netherlands\\
\Curtin{}International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia\\
\UWisc{}Department of Physics, University of Wisconsin--Milwaukee, Milwaukee, WI 53201, USA\\
\UW{}Department of Physics, University of Washington, Seattle, WA 98195, USA\\
\CASS{}CSIRO Astronomy and Space Science, Marsfield, NSW 2122, Australia\\
\SKASA{}Square Kilometre Array South Africa (SKA SA), Cape Town, 7405, South Africa\\
\ASU{}School of Earth and Space Exploration, Arizona State University, Tempe, AZ 85287, USA\\
\USydney{}Sydney Institute for Astronomy, School of Physics, University of Sydney, NSW 2006, Australia\\
\Haystack{}MIT Haystack Observatory, Westford, MA 01886, USA\\
\RRI{}Raman Research Institute, Bangalore 560080, India\\
\MIT{}Kavli Institute for Astrophysics and Space Research, Massachusetts Institute of Technology, Cambridge, MA 02139, USA\\
\UWA{}University of Western Australia, Crawley, WA 6009, Perth, Australia\\
\CfA{}Harvard-Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA\\
\UMelbourne{}School of Physics, University of Melbourne, Parkville, VIC 3010, Australia\\
\Victoria{}School of Chemical \& Physical Sciences, Victoria University of Wellington, Wellington 6140, New Zealand\\
\Tata{}National Centre for Radio Astrophysics, Tata Institute for Fundamental Research, Pune 411007, India
%\UMichigan{}University of Michigan, Ann Harbor, MI 48109, USA
}
\jid{PASA}
\doi{10.1017/pas.\the\year.xxx}
\jyear{\the\year}
%Citations may be made using the natbib commands \citet{},\citep{} etc.
\usepackage[authoryear]{natbib}
\bibpunct{(}{)}{;}{a}{}{,}
\setlength{\bibsep}{0.3mm}

\begin{document}
\begin{abstract}
The Murchison Widefield Array (MWA) is a new low-frequency interferometric radio telescope built in Western Australia at one of the locations of the future Square Kilometre Array (SKA). We describe the automated RFI detection strategy implemented for the MWA, which is based on the \textsc{aoflagger} platform, and present 72--231-MHz RFI statistics from 10 observing nights. RFI detection removes 1.1\% of the data. RFI from digital TV (DTV) is observed 3\% of the time due to occasional ionospheric or atmospheric propagation. After RFI detection, almost all data can be calibrated and imaged without further RFI mitigation efforts, including observations within the FM and DTV bands. The results are compared to a previously published LOFAR RFI survey. The remote location of the MWA has resulted in a substantially cleaner RFI environment compared to LOFAR's radio environment, but adequate detection of RFI is still required before analysis of data. We include specific recommendations designed to make the SKA more robust to RFI, including: the availability of sufficient computing power for RFI detection; analogue-digital converters with sufficient dynamic range; a smooth band-pass response; and the possibility for RFI detection at high time and frequency resolution (second and kHz-scale respectively).
\end{abstract}

@@

When time or frequency averaging is requested, \textsc{cotter} averages samples together that have not been flagged. When all input samples are flagged, the average of all input visibilities is stored into the output sample, and the output sample is flagged. This method makes it possible to superficially analyse flagged samples in the output, even though information is lost in the averaging. \textsc{cotter} stores a weight for each visibility in the output file, which is scaled to the number of unflagged input samples that were used for the output sample. Because of this, when no averaging is requested, the output is 50\% larger than the input (one extra float per complex float visibility). In practice, most observations are recorded at a resolution of 0.5 s and 40 kHz, and averaged to a resolution of 2--4 s and 40--80 kHz to reduce the data volume. This does not cause any significant time or bandwidth decorrelation up to the first beam null. \textsc{cotter} performs the phase shifting and cable delay corrections before averaging, and recalculates the $u,v,w$ values for the central time and frequency of the output sample. This helps to prevent time/frequency decorrelation.

@@

\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-stddev-per-set}\vspace{-1cm}
\caption{Observed visibility RMS in the high-resolution (40~kHz) channels for each observation. This is the residual noise in the visibilities after RFI detection and flagging. A few channels in the FM bands show a deviating standard deviation, and observation `GLEAM 2014-03-17' shows DTV contamination around 180~MHz. The variability over nights is caused by the different celestial observing times and pointing directions, and therefore different apparent brightness of e.g. the Galaxy.}
\label{fig:stddev-per-set}
\end{center}
\end{figure*}

@@

After RFI detection, power spectra from both MWA and LOFAR show smooth curves without significant leakage, and the data are clean enough for further processing and imaging. From the difference between 1.65\% or 3.18\% loss of data at the MWA and LOFAR sites respectively, it is clear that the impact of RFI is smaller for the MWA, but RFI mitigation is still required with similar per-sample computational requirements as for LOFAR. Both RFI occupancy levels are so small that the RFI contamination in the 115--163~MHz range should not be a consideration in selecting a site. However, a benefit of building the MWA at a remote radio-quiet site, is that the MWA can observe in the 88--108~MHz FM-station and 174-230 MHz DTV bands. Initial MWA experiments have confirmed the availability of these bands for science \citep{mckinley-moon-2013,mwacs-2014}. Moreover, for both LOFAR and MWA, a primary science driver is the detection of signals from the EoR. The required integration time for such a detection of approximately a hundred nights has currently not been achieved, and it could be that RFI becomes problematic when reaching lower noise levels. Initial results with LOFAR predict that RFI will not prevent such a detection \citep{ncp-eor-yatawatta, offringa-rfi-distributions}.

>>>>

\author[A.~R.~Offringa et al.]{A.~R.~Offringa\ANU$^,$\CAASTRO$^,$\ASTRON\thanks{Corresponding author. E-mail: \url{offringa@gmail.com}},
R.~B.~Wayth\Curtin$^,$\CAASTRO,
N.~Hurley-Walker\Curtin,
D.~L.~Kaplan\UWisc,
N.~Barry\UW,
A.~P.~Beardsley\UW,
M.~E.~Bell\CASS,
G.~Bernardi\SKASA,
J.~D.~Bowman\ASU, 
F.~Briggs\ANU,
J.~R.~Callingham\USydney$^,$\CAASTRO,
R.~J.~Cappallo\Haystack, 
P.~Carroll\UW,
%A.~de~Oliveira-Costa
A.~A.~Deshpande\RRI, 
J.~S.~Dillon\MIT,
K.~S.~Dwarakanath\RRI,
A.~Ewall-Wice\MIT,
L.~Feng\MIT,
B.-Q.~For\UWA,
B.~M.~Gaensler\USydney$^,$\CAASTRO, 
L.~J.~Greenhill\CfA,
P.~Hancock\Curtin,
K.~Han-Seek\UMelbourne,
B.~J.~Hazelton\UW,
J.~N.~Hewitt\MIT,
L.~Hindson\Victoria,
D.~C.~Jacobs\ASU, 
M.~Johnston-Hollitt\Victoria,
A.~Kapinska\UWA$^,$\CAASTRO,
P.~Kittiwisit\ASU,
E.~Lenc\USydney$^,$\CAASTRO,
J.~Line\UMelbourne,
A.~Loeb\CfA,
C.~J.~Lonsdale\Haystack, 
B.~McKinley\ANU$^,$\CAASTRO,
S.~R.~McWhirter\Haystack,
D.~A.~Mitchell\CASS$^,$\CAASTRO, 
M.~F.~Morales\UW, 
E.~Morgan\MIT, 
J.~Morgan\Curtin,
A.~R.~Neben\MIT,
D.~Oberoi\Tata, 
S.~M.~Ord\Curtin$^,$\CAASTRO,
S.~Paul\RRI,
B.~Pindor\UMelbourne,
J.~C.~Pober\UW,
T.~Prabu\RRI, 
P.~Procopio\UMelbourne,
J.~Riding\UMelbourne,  
N.~Udaya~Shankar\RRI, 
S.~Sethi\RRI,
K.~S.~Srivani\RRI,
L.~Staveley-Smith\UWA,
R.~Subrahmanyan\RRI$^,$\CAASTRO, 
I.~S.~Sullivan\UW,
M.~Tegmark\MIT,
N.~Thyagarajan\ASU,
S.~J.~Tingay\Curtin$^,$\CAASTRO, 
C.~M.~Trott\Curtin$^,$\CAASTRO,
R.~L.~Webster\UMelbourne$^,$\CAASTRO, 
A.~Williams\Curtin, 
C.~L.~Williams\MIT,
C.~Wu\UWA,
J.~S.~Wyithe\UMelbourne,
Q. Zheng\Victoria
\\
\ANU{}Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
%\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics (CAASTRO) \\
\ASTRON{}Netherlands Institute for Radio Astronomy (ASTRON), PO Box 2, 7990 AA Dwingeloo, The Netherlands\\
\Curtin{}International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia\\
\UWisc{}Department of Physics, University of Wisconsin--Milwaukee, Milwaukee, WI 53201, USA\\
\UW{}Department of Physics, University of Washington, Seattle, WA 98195, USA\\
\CASS{}CSIRO Astronomy and Space Science, Marsfield, NSW 2122, Australia\\
\SKASA{}Square Kilometre Array South Africa (SKA SA), Cape Town, 7405, South Africa\\
\ASU{}School of Earth and Space Exploration, Arizona State University, Tempe, AZ 85287, USA\\
\USydney{}Sydney Institute for Astronomy, School of Physics, University of Sydney, NSW 2006, Australia\\
\Haystack{}MIT Haystack Observatory, Westford, MA 01886, USA\\
\RRI{}Raman Research Institute, Bangalore 560080, India\\
\MIT{}Kavli Institute for Astrophysics and Space Research, Massachusetts Institute of Technology, Cambridge, MA 02139, USA\\
\UWA{}University of Western Australia, Crawley, WA 6009, Perth, Australia\\
\CfA{}Harvard-Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA\\
\UMelbourne{}School of Physics, University of Melbourne, Parkville, VIC 3010, Australia\\
\Victoria{}School of Chemical \& Physical Sciences, Victoria University of Wellington, Wellington 6140, New Zealand\\
\Tata{}National Centre for Radio Astrophysics, Tata Institute for Fundamental Research, Pune 411007, India
%\UMichigan{}University of Michigan, Ann Harbor, MI 48109, USA
}
\jid{PASA}
\doi{10.1017/pas.\the\year.xxx}
\jyear{\the\year}
%Citations may be made using the natbib commands \citet{},\citep{} etc.
\usepackage[authoryear]{natbib}
\bibpunct{(}{)}{;}{a}{}{,}
\setlength{\bibsep}{0.3mm}

\begin{document}
\begin{abstract}
The Murchison Widefield Array (MWA) is a new low-frequency interferometric radio telescope built in Western Australia at one of the locations of the future Square Kilometre Array (SKA). We describe the automated RFI detection strategy implemented for the MWA, which is based on the \textsc{aoflagger} platform, and present 72--231-MHz RFI statistics from 10 observing nights. RFI detection removes 1.1\% of the data. RFI from digital TV (DTV) is observed 3\% of the time due to occasional ionospheric or atmospheric propagation. After RFI detection, almost all data can be calibrated and imaged without further RFI mitigation efforts, including observations within the FM and DTV bands. The results are compared to a previously published LOFAR RFI survey. The remote location of the MWA has resulted in a substantially cleaner RFI environment compared to LOFAR's radio environment, but adequate detection of RFI is still required before data can be analysed. We include specific recommendations designed to make the SKA more robust to RFI, including: the availability of sufficient computing power for RFI detection; analogue-digital converters with sufficient dynamic range; a smooth band-pass response; and the possibility for RFI detection at high time and frequency resolution (second and kHz-scale respectively).
\end{abstract}

@@

When time or frequency averaging is requested, \textsc{cotter} averages samples together that have not been flagged. When all input samples are flagged, the average of all input visibilities is stored into the output sample, and the output sample is flagged. This method makes it possible to superficially analyse flagged samples in the output, even though information is lost in the averaging. \textsc{cotter} stores a weight for each visibility in the output file, which is scaled to the number of unflagged input samples that were used for the output sample. Because of this, when no averaging is requested, the output is 50\% larger than the input (one extra float per complex float visibility). In practice, most MWA observations are recorded at a resolution of 0.5~s and 40~kHz, and averaged to a resolution of 2--4 s and 40--80 kHz to reduce the data volume. This does not cause any significant time or bandwidth decorrelation up to the first beam null for MWA data. \textsc{cotter} performs the phase shifting and cable delay corrections before averaging, and recalculates the $u,v,w$ values for the central time and frequency of the output sample. This helps to prevent time/frequency decorrelation.

@@

\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-stddev-per-set}\vspace{-1cm}
\caption{Observed visibility RMS in the high-resolution (40~kHz) channels for each observation. This is the residual noise in the visibilities after RFI excision. A few channels in the FM bands show a deviating standard deviation, and observation `GLEAM 2014-03-17' shows DTV contamination around 180~MHz. The variability over nights is caused by the different celestial observing times and pointing directions, and therefore the difference in apparent brightness of e.g. the Galaxy.}
\label{fig:stddev-per-set}
\end{center}
\end{figure*}

@@

After RFI detection, power spectra from both MWA and LOFAR show smooth curves without significant leakage, and the data are clean enough for further processing and imaging. From the difference between 1.65\% or 3.18\% loss of data at the MWA and LOFAR sites respectively, it is clear that the impact of RFI is smaller for the MWA, but RFI mitigation is still required with similar per-sample computational requirements as for LOFAR. When comparing the 115--163~MHz range, the RFI occupancy levels are small and the RFI flagging is effective for both sites. Therefore, the MWA's larger radio-quiet zone does not provide a significant benefit at these frequencies. However, a benefit of the remote radio-quiet site, is that the MWA can observe in the 88--108~MHz FM-station and 174-230 MHz DTV bands. Initial MWA experiments have confirmed the availability of these bands for science \citep{mckinley-moon-2013,mwacs-2014}. Moreover, for both LOFAR and MWA, a primary science driver is the detection of signals from the EoR. The required integration time for such a detection of approximately a hundred nights has currently not been achieved, and it could be that RFI becomes problematic when reaching lower noise levels. Initial results with LOFAR predict that RFI will not prevent such a detection \citep{ncp-eor-yatawatta, offringa-rfi-distributions}.
